BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-03-09 Paper/D5.5/D5.5.tex
Updated deliverable.

@@

\begin{executive}
This document describes BOAST a metaprogramming framework to produce portable and efficient computing kernels for HPC application.
\end{executive}

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources.

Portability efforts are often lost when migrating to a new architecture, or code
lose maintainability because several versions of the code coexist, usually with
a lot of duplication.

@@

Genericity of HPC codes is often limited. Producing generic code in FORTRAN
90/95 is difficult as the language is not really  fit for it.

@@

Arrays which correspond to output parameters will be modified during the
execution of the kernel so results can be checked. This allows for easy non
regression testing. The run method also returns information (and result for
kernels that are functions) about the run. For instance the run time of the
kernel is one such information, it is obtained using the system wide realtime
clock. Other probes can be inserted at compile time if needed.

@@

In the Kohn-Sham (KS) formulation of DFT, the electrons are associated to
wavefunctions (orbitals), which are represented arrays of floating point
numbers. In wavelets formalism, the operators are written via convolutions with
short, separable filters.  The detailed description of how these operations are
defined is beyond the scope of this paper and can be found in the BigDFT
reference paper \cite{bigdft}.

@@

@@

>>>>

\lstdefinestyle{custombash}{
  language=Bash,
  basicstyle=\scriptsize,
  backgroundcolor=\color{white},
  frame=single,
  captionpos=b
}

@@

\begin{executive}
This document describes BOAST, a metaprogramming framework to produce portable and efficient computing kernels for HPC application.
\end{executive}

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources. Nonetheless it is a very important aspect of the
Mont-Blanc project. Indeed, for the project, more than 10 applications were
selected to be ported and optimized for the prototype platform.

Unfortunately, portability efforts are often lost when migrating to a new
architecture. Worse, code may lose maintainability because several versions of
some functionalities coexist, usually with a lot of duplication.

@@

Genericity of HPC codes is often limited. One of the reason is that producing
generic code in FORTRAN 90/95 is difficult as the language is not really fit
for it. Sometimes, adding genericity will degrade performance as optimization
opportunities that come from over-specification are lost.

@@

Arrays which correspond to output parameters will be modified during the
execution of the kernel so results can be checked. This allows for easy non
regression testing. The run method also returns information (and result for
kernels that are functions as well as output scalars) about the run. For
instance the run time of the kernel is one such information, it is obtained
using the system wide realtime clock. Other probes can be inserted at compile
time if needed.

PAPI~\cite{mucci1999papi} is supported and different counters can be specified
during each execution of a kernel.

@@

In the Kohn-Sham (KS) formulation of DFT, the electrons are associated to
wavefunctions (orbitals), which are represented arrays of floating point
numbers. In wavelets formalism, the operators are written via convolutions with
short, separable filters.  The detailed description of how these operations are
defined is beyond the scope of this report and can be found in the BigDFT
reference paper \cite{bigdft}.

@@

In this report we presented the BOAST infrastructure as well as its
application to two use cases from the Mont-Blanc project. Results are
encouraging as BOAST proved to be a powerful and flexible tool.

@@

\section*{Annexes}

This section will present some simple examples to familiarize the user with
BOAST. More samples can be found in the git repository.

\subsection*{Installation}

BOAST is ruby based, so ruby needs to be installed on the machine.
Installation of boast can be done using the ruby built-in package manager:
\textit{gem}. See Listing~\ref{lst:BOAST_install} for reference.

\lstset{style=custombash,
        caption={BOAST installation}, 
        label={lst:BOAST_install},
        numbers=left }
\begin{lstlisting}
sudo apt-get install ruby ruby-dev
gem install --user-install BOAST
\end{lstlisting}

\subsection*{Variable and Procedure Declaration}

The following samples are presented using \textit{irb} ruby interactive
interpreter. It can be launched using the \textit{irb} command in a terminal.
Listing~\ref{lst:variable_declaration} shows the declaration of two variables
of different kind.

\lstset{style=BOAST,
        caption={Variable Declaration},
        label={lst:variable_declaration}}
\begin{lstlisting}
irb(main):001:0> require 'BOAST'
=> true
irb(main):002:0> a = BOAST::Int "a"
=> a
irb(main):003:0> b = BOAST::Real "b"
=> b
irb(main):004:0> BOAST::decl a, b
integer(kind=4) :: a
real(kind=8) :: b
=> [a, b]
\end{lstlisting}

Listing~\ref{lst:procedure_declaration} shows the declaration of a procedure
using the two previous variables as parameters. For clarity irb echoes have
been suppressed.

\lstset{style=BOAST,
        caption={Procedure Declaration},
        label={lst:procedure_declaration}}
\begin{lstlisting}
005:0> p = BOAST::Procedure( "test_proc", [a,b] )
006:0> BOAST::opn p
SUBROUTINE test_proc(a, b)
  integer, parameter :: wp=kind(1.0d0)
  integer(kind=4) :: a
  real(kind=8) :: b
007:0> BOAST::close p
END SUBROUTINE test_proc
\end{lstlisting}

\subsection*{Switching Language}

Listing~\ref{lst:switching_language} shows how to switch BOAST to C. Available
languages are \textit{FORTRAN}, \textit{C}, \textit{CUDA} and \textit{CL}.

\lstset{style=BOAST,
        caption={Switching Language},
        label={lst:switching_language}}
\begin{lstlisting}
008:0> BOAST::lang = BOAST::C
009:0> BOAST::opn p
void test_proc(int32_t a, double b){
010:0> BOAST::close p
}
\end{lstlisting}

\subsection*{Defining a Complete Procedure}

Listing~\ref{lst:complete_procedure} shows how to define a procedure and the
associated code. Note that here the parameters of the procedure have been
associated a direction: one, \textit{a}, is an input parameter while the other
is an output parameter.

\lstset{style=BOAST,
        caption={Complete Procedure},
        label={lst:complete_procedure}}
\begin{lstlisting}
011:0> BOAST::lang = BOAST::FORTRAN
012:0> a = BOAST::Real( "a", :dir => :in)
013:0> b = BOAST::Real( "b", :dir => :out)
014:0> p = BOAST::Procedure( "plus_two", [a,b] ) {
015:1*   BOAST::pr b === a + 2
016:1> }
017:0> BOAST::pr p
SUBROUTINE plus_two(a, b)
  integer, parameter :: wp=kind(1.0d0)
  real(kind=8), intent(in) :: a
  real(kind=8), intent(out) :: b
  b = a + 2
END SUBROUTINE plus_two
018:0> BOAST::lang = BOAST::C
019:0> BOAST::pr p
void plus_two(const double a, double * b){
  (*b) = a + 2;
}
\end{lstlisting}

\subsection*{Creating, Building and Running a Computing Kernel}

Listing~\ref{lst:computing_kernel} shows how to create a Computing kernel
(\textit{CKernel}) and build it. Once a computing kernel is instancied the
output of BOAST will be redirected to the computing kernel source code. Line 4
sets the entry point of the computing kernel to the procedure we just defined.
By default compilation commands are not shown unless an error occurs. This
behavior can be change by switching to verbose mode.

When running the kernel all the arguments have to be specified. Running a
kernel returns a hash table containing information about the procedure
execution. In this simple case two information are returned, first the value
of the output parameter \textit{b} and second the time the kernel execution
took.

\lstset{style=BOAST,
        caption={Computing Kernel},
        breaklines=true,
        breakautoindent=false,
        label={lst:computing_kernel}}
\begin{lstlisting}
020:0> BOAST::lang = BOAST::FORTRAN
021:0> k = BOAST::CKernel::new
022:0> BOAST::pr p
023:0> k.procedure = p
024:0> puts k
SUBROUTINE plus_two(a, b)
  integer, parameter :: wp=kind(1.0d0)
  real(kind=8), intent(in) :: a
  real(kind=8), intent(out) :: b
  b = a + 2
END SUBROUTINE plus_two
025:0> k.build
026:0> BOAST::verbose = true
027:0> k.build
gcc -O2 -Wall -fPIC -I/usr/lib/x86_64-linux-gnu/ruby/2.1.0 -I/usr/include/ruby-2.1.0 -I/usr/include/ruby-2.1.0/x86_64-linux-gnu -I/usr/include/x86_64-linux-gnu/ruby-2.1.0 -I/var/lib/gems/2.1.0/gems/narray-0.6.1.1 -DHAVE_NARRAY_H -c -o /tmp/Mod_plus_two20150309_4611_5a129k.o /tmp/Mod_plus_two20150309_4611_5a129k.c
gfortran -O2 -Wall -fPIC -c -o /tmp/plus_two20150309-4611-5a129k.o /tmp/plus_two20150309-4611-5a129k.f90
gcc -shared -o /tmp/Mod_plus_two20150309_4611_5a129k.so /tmp/Mod_plus_two20150309_4611_5a129k.o /tmp/plus_two20150309-4611-5a129k.o -Wl,-Bsymbolic-functions -Wl,-z,relro -rdynamic -Wl,-export-dynamic -L/usr/lib -lruby-2.1 -lrt
028:0> r = k.run(5,0)
029:0> puts r
{:reference_return=>{:b=>7.0}, :duration=>5.84e-07}
\end{lstlisting}

\subsection*{Using Arrays in Procedures}

Most computing kernels don't work on scalar values but rather on array of
data. Listing~\ref{lst:arrays} shows how to use arrays in computing kernels.
In this case we place ourselves in BOAST namespace to reduce the syntax
overhead. Variables \textit{a} and \textit{b} are one-dimensional arrays of
size \textit{n}. Arrays in BOAST start at index 1 unless specified otherwise.
For instance \lstinline!Dim(0,n-1)! would have created a dimension starting at
0. Array bounds can also be negative and several dimensions can be specified
to obtain muti-dimensional arrays.
For self contained procedures/kernels one can sue the shortcut written on
line~13 to create a CKernel object. As we are not specifying build options the
build command can also be ommited and will be automatically called when
running the kernel the first time. Lines~17 to 19 are used to check the result
of the kernel.

\lstset{style=BOAST,
        caption={Array Usage},
        label={lst:arrays}}
\begin{lstlisting}
001:0> require 'BOAST'
002:0> require 'narray'
003:0> include BOAST
004:0> n = Int(  "n", :dir => :in )
005:0> a = Real( "a", :dir => :in,  :dim => [Dim(n)] )
006:0> b = Real( "b", :dir => :out, :dim => [Dim(n)] )
007:0> p = Procedure( "plus_two", [n, a, b] ) {
008:1*   decl i = Int( "i" )
009:1>   pr For( i, 1, n ) {
010:2*     pr b[i] === a[i] + 2.0
011:2>   }
012:1> }
013:0> k = p.ckernel
014:0> input  = NArray.float(1024).random
015:0> output = NArray.float(1024)
016:0> k.run(input.length, input, output)
017:0> (output - input).each { |val|
018:1*   raise "Error!" if (val-2).abs > 1e-15
019:1> }
020:0> stats = k.run(input.length, input, output)
021:0> puts "Success, duration: #{stats[:duration]} s"
Success, duration: 3.79e-06 s
\end{lstlisting}

\subsection*{The Canonical Case: Vector Addition}

Listing~\ref{lst:vector_add_decl} shows the addition of two vector in a third
one. Here BOAST is configured to have arrays starting at 0 and to use single
precision reals by default (Lines~5 and 6).  The kernel declaration is
encapsulated inside a method to avoid cluttering the global namespace. Line 15
the expression \lstinline!c[i] === a[i] + b[i]! is stored inside a variable
\textit{expr} for later use. Lines~16 to 23 show that the kernel differs
depending on the target language, in CUDA and OpenCL each thread will process
one element.

\lstset{style=BOAST,
        caption={Vector Addition Declaration},
        label={lst:vector_add_decl}}
\begin{lstlisting}
require 'narray'
require 'BOAST'
include BOAST

set_array_start(0)
set_default_real_size(4)

def vector_add
  n = Int("n",:dir => :in)
  a = Real("a",:dir => :in, :dim => [ Dim(n)] )
  b = Real("b",:dir => :in, :dim => [ Dim(n)] )
  c = Real("c",:dir => :out, :dim => [ Dim(n)] )
  p = Procedure("vector_add", [n,a,b,c]) {
    decl i = Int("i")
    expr = c[i] === a[i] + b[i]
    if (get_lang == CL or get_lang == CUDA) then
      pr i === get_global_id(0)
      pr expr
    else
      pr For(i,0,n-1) {
        pr expr
      }
    end
  }
  return p.ckernel
end
\end{lstlisting}

Listing~\ref{lst:vector_add_run} shows the a way to check the validity of the
previous kernel over the available range of languages. The options that are
passed to run are only relevant for GPU languages and are thus ignored in
FORTRAN and C (Line~16). Success is only printed if results are validated,
else an exception is raised (Lines~17 to 20).

\lstset{style=BOAST,
        caption={Vector Addition Test},
        label={lst:vector_add_run}}
\begin{lstlisting}
n = 1024*1024
a = NArray.sfloat(n).random
b = NArray.sfloat(n).random
c = NArray.sfloat(n)

epsilon = 10e-15

c_ref = a + b

[:FORTRAN, :C, :CL, :CUDA].each { |l|
  set_lang( BOAST.const_get(l)  )
  puts "#{l}:"
  k = vector_add
  puts k.print
  c.random!
  k.run(n, a, b, c, :global_work_size => [n,1,1], :local_work_size => [32,1,1])
  diff = (c_ref - c).abs
  diff.each { |elem|
    raise "Warning: residue too big: #{elem}" if elem > epsilon
  }
}
puts "Success!"
\end{lstlisting}
